---
title: "Boosting II"
output: 
  learnr::tutorial:
    progressive: true
runtime: shiny_prerendered
tutorial:
  id: "boosting-2"
  version: 0.5
---

## Setup

```{r results='hide', message=FALSE, warning=FALSE}
# devtools::install_github("AppliedDataSciencePartners/xgboostExplainer")
library(caret)
library(xgboost)
library(mboost)
library(ranger)
library(xgboostExplainer)
```

## Data

In this notebook we (again) use the drug consumption data. The data contains records for 1885 respondents with personality measurements (e.g. Big-5), level of education, age, gender, country of residence and ethnicity as features. In addition, information on the usage of 18 drugs is included. 

Source: https://archive.ics.uci.edu/ml/datasets/Drug+consumption+%28quantified%29

```{r}
drugs <- read.csv("drug_consumption.data", header = FALSE)
names(drugs) <- c("ID", "Age", "Gender", "Education", "Country", "Ethnicity", "Neuroticism", "Extraversion", "Openness", "Agreeableness", "Conscientiousness", "Impulsive", "SS", "Alcohol", "Amphet", "Amyl", "Benzos", "Caff", "Cannabis", "Choc", "Coke", "Crack", "Ecstasy", "Heroin", "Ketamine", "Legalh", "LSD", "Meth", "Mushrooms", "Nicotine", "Semer", "VSA")
drugs$Age <- as.factor(drugs$Age)
drugs$Gender <- as.factor(drugs$Gender)
drugs$Education <- as.factor(drugs$Education)
drugs$Country <- as.factor(drugs$Country)
drugs$Ethnicity <- as.factor(drugs$Ethnicity)
```

First, we build a dummy variable on LSD usage as our outcome of interest.

```{r}
drugs$D_LSD <- "LSD"
drugs$D_LSD[drugs$LSD == "CL0"] <- "no_LSD"
drugs$D_LSD <- as.factor(drugs$D_LSD)
summary(drugs$D_LSD)
```

Then we split the data into a training and a test part, using `createDataPartition` from `caret`.

```{r}
set.seed(9453)
inTrain <- createDataPartition(drugs$D_LSD, 
                               p = .8, 
                               list = FALSE, 
                               times = 1)

drugs_train <- drugs[inTrain,]
drugs_test <- drugs[-inTrain,]
```

## XGBoost

We start with `xgboost` as our ML method, which is a efficient and competitive Boosting implementation. We first specify the resampling method for tuning with `trainControl()`.

```{r}
ctrl  <- trainControl(method = "cv",
                      number = 5,
                      summaryFunction = twoClassSummary,
                      verboseIter = TRUE,
                      classProbs = TRUE)
```

Using XGBoost requires to take care of many tuning parameters (see `?xgboost`). For this example, we fix most of them. 

```{r}
grid <- expand.grid(max_depth = c(1, 3, 5),
                    nrounds = c(500, 1000, 1500),
                    eta = c(0.05, 0.01, 0.005),
                    min_child_weight = 10,
                    subsample = 0.7,
                    gamma = 0,
                    colsample_bytree = 1)
```

Print the tuning grid.

```{r}
grid
```

The grid object can now be used in the `train` function to guide the tuning process.

```{r}
set.seed(8303)
xgb <- train(D_LSD ~ Age + Gender + Education + Neuroticism + Extraversion +
             Openness + Agreeableness + Conscientiousness + Impulsive + SS,
             data = drugs_train,
             method = "xgbTree",
             trControl = ctrl,
             tuneGrid = grid,
             metric = "ROC")
```

Plot the tuning results. 

```{r, fig.align="center"}
plot(xgb)
```

## mboost

Here we use Model-Based Boosting as an alternative boosting approach. It has considerably fewer tuning parameters as XGBoost -- in fact, via `caret` we primarily have to take care of the number of boosting iterations.

```{r}
grid <- expand.grid(mstop = c(50, 100, 150, 200, 250, 500),
                    prune = 'no')
```

Now we run the tuning process with `train()`, using `glmboost` as the prediction method.

```{r}
set.seed(8303)
mboost <- train(D_LSD ~ Age + Gender + Education + Neuroticism + Extraversion +
                Openness + Agreeableness + Conscientiousness + Impulsive + SS,
                data = drugs_train,
                method = "glmboost",
                trControl = ctrl,
                tuneGrid = grid,
                metric = "ROC")
```

Print the tuning results.

```{r}
mboost
```

With `mboost` it is possible to access model "coefficients".

```{r}
coef(mboost$finalModel)
```

## Random Forest (and random search)

This section exemplifies the usage of random search (as opposed to grid search). In `caret`, random search is available as an option of the `trainControl()` function.

```{r}
ctrl  <- trainControl(method = "cv",
                      number = 5,
                      summaryFunction = twoClassSummary,
                      classProbs = TRUE,
                      search = "random")
```

Note that now we don't set up a tuning grid, but specify the number of random picks of tuning parameters settings via `tuneLength` in `train()`. We use `ranger` as our ML method, which implements random forests and extremely randomized trees.

```{r}
set.seed(8303)
ranger <- train(D_LSD ~ Age + Gender + Education + Neuroticism + Extraversion + 
                Openness + Agreeableness + Conscientiousness + Impulsive + SS,
                data = drugs_train,
                method = "ranger",
                trControl = ctrl,
                tuneLength = 10,
                metric = "ROC")
```

Print the tuning results (and show the randomly selected try-out values).

```{r}
ranger
```

## Comparison

This section shows some options for comparing the CV results of different methods. After we ran a bunch of models, we can use carets `resamples()` function to gather the cross-validation results from all of them.

```{r}
resamps <- resamples(list(XGBoost = xgb,
                          mboost = mboost,
                          RF = ranger))
```

This object can now be used for comparing these models with respect to their performance, based on Cross-Validation in the training set.

```{r}
summary(resamps)
```

We can also plot this information.

```{r}
bwplot(resamps)
```

Another option is to compare the resampling distributions with scatterplots.

```{r}
splom(resamps)
```

## XGBoost Explainer

Finally, we take a look at the `xgboostExplainer` package. Currently it does not work with XGBoost models that were build with `caret`, i.e. we have to use the `xgboost` package to re-build our XGBoost model. We first need a model matrix.

```{r}
d_matrix <- model.matrix(D_LSD ~ Age + Gender + Education + Neuroticism + Extraversion + Openness + Agreeableness + Conscientiousness + Impulsive + SS, data = drugs)[,-1]
```

This matrix can be used to create the data objects needed for `xgboost`.

```{r}
xgb_train_data <- xgb.DMatrix(d_matrix[inTrain,], label = ifelse(drugs_train$D_LSD == "LSD", 1, 0))
xgb_test_data <- xgb.DMatrix(d_matrix[-inTrain,])
```

On this basis, the XGBoost model can be fitted. 

```{r}
xgb_model <- xgboost(param = list(objective = "binary:logistic"), 
                     data = xgb_train_data, 
                     nrounds = 1000,
                     max_depth = 1,
                     eta = 0.01,
                     gamma = 0,
                     colsample_bytree = 1,
                     min_child_weight = 10,
                     subsample = 0.7,
                     verbose = 0)
```

This `xgb_model` object can now be used to build the explainer object, which in turn is needed to `explainPredictions()`.

```{r, message = FALSE}
explainer <- buildExplainer(xgb_model, xgb_train_data, type = "binary")
train_expl <- explainPredictions(xgb_model, explainer, xgb_test_data)
train_expl
```

As the final step, predictions for individual (test) observations can be explained graphically with `showWaterfall()`.

```{r, message = FALSE}
showWaterfall(xgb_model, explainer, xgb_test_data, d_matrix[-inTrain,], 1, type = "binary")
showWaterfall(xgb_model, explainer, xgb_test_data, d_matrix[-inTrain,], 2, type = "binary")
```

## References

* https://github.com/AppliedDataSciencePartners/xgboostExplainer
