---
title: "Interpretable Machine Learning"
subtitle: "An Example using Random Forests"
output: 
  learnr::tutorial:
    progressive: true
runtime: shiny_prerendered
tutorial:
  id: "knn"
  version: 0.5
---

## Setup

```{r results='hide', message=FALSE, warning=FALSE}
library(caret)
library(ranger)
library(pdp)
library(iml)
library(DALEX)
library(mlforsocialscience)
```

### Data

In this notebook, we use the Communities and Crime data from the UCI ML repository, which includes information about communities in the US. "The data combines socio-economic data from the 1990 US Census, law enforcement data from the 1990 US LEMAS survey, and crime data from the 1995 FBI UCR"

Source: https://archive.ics.uci.edu/ml/datasets/Communities+and+Crime

First, some data prep. We will use the `crime` data which has been included as part of the `mlforsocialscience` package.

```{r}
data(crime)
```

Build a binary outcome variable.

```{r}
quantile(crime$ViolentCrimesPerPop)
crime$D_crime <- ifelse(crime$ViolentCrimesPerPop > 0.33, "high", "not_high")
crime$D_crime <- as.factor(crime$D_crime)
table(crime$D_crime)
```

Final clean up.

```{r}
crime$state <- NULL
crime$communityname <- NULL
crime$fold <- NULL
crime$ViolentCrimesPerPop <- NULL
```

Split the data into a training and a test part. This time we use `createDataPartition()` from `caret`, which samples within the levels of the outcome variable when splitting the data.

```{r}
set.seed(9282)
inTrain <- createDataPartition(crime$D_crime, 
                               p = .8, 
                               list = FALSE, 
                               times = 1)

crime_train <- crime[inTrain,]
crime_test <- crime[-inTrain,]
```

## RF and Extra-Trees

We start by building random forests and Extra-Trees using the `caret` package. We first specify our evaluation method -- e.g., 5-fold cross-validation.

```{r}
ctrl  <- trainControl(method = "cv",
                      number = 5,
                      summaryFunction = twoClassSummary,
                      classProbs = TRUE)
```

Next, a grid object can be used to specify a set of try-out values. For random forest and Extra-Trees, we have to care about `mtry`, i.e. the number of features to sample at each split point. Furthermore, the `ranger` package treats the type of forest as a tuning parameter.

```{r}
grid <- expand.grid(mtry = c(sqrt(ncol(crime_train)),
                             log(ncol(crime_train))),
                    splitrule = c("gini", "extratrees"),
                    min.node.size = 10)
grid
```

The grid object can be passed on to `train()`, along with the specification of the model and the method (`ranger`).

```{r}
set.seed(9385)
rf <- train(D_crime ~ .,
            data = crime_train,
            method = "ranger",
            trControl = ctrl,
            tuneGrid = grid,
            metric = "ROC",
            importance = "impurity")
```

Calling the ranger object lists the results of the tuning process.

```{r}
rf
```

With ranger, `treeInfo()` can be used to print individual tree results. 

```{r}
treeInfo(rf$finalModel, tree = 1)
```

## Variable Importance, PDP and ICE

We start with plotting variable importances to get a general idea of which variables were most effective in reducing impurity when building the random forest.

```{r}
plot(varImp(rf), top = 10)
```

Partial dependence plots can be useful in order to see how the features are related to the outcome according to the fitted model. With the `pdp` package, we start by running the `partial()` function with the variables of interest.

```{r}
pdp1 <- partial(rf, pred.var = "PctIlleg", 
                type = "classification", 
                which.class = 1, prob = T)
pdp2 <- partial(rf, pred.var = "racePctWhite", 
                type = "classification", 
                which.class = 1, prob = T)
```

The actual plots can be created with `plotPartial()`.

```{r}
p1 <- plotPartial(pdp1, rug = T, train = crime_train)
p2 <- plotPartial(pdp2, rug = T, train = crime_train)
grid.arrange(p1, p2, ncol = 2)
```

We can also consider multiple predictors jointly.

```{r, message = FALSE, results = "hide"}
pdp3 <- partial(rf, pred.var = c("PctIlleg", "racePctWhite"), 
                type = "classification", 
                which.class = 1, prob = T, 
                grid.resolution = 20, progress = "text")
plotPartial(pdp3, levelplot = F, drape = T, colorkey = F, 
            screen = list(z = 45, x = -60))
```

A centered ICE plot example with the `pdp` package.

```{r}
pdp4 <- partial(rf, pred.var = "PctIlleg", 
                type = "classification", 
                which.class = 1, prob = T, 
                ice = TRUE, center = T)
plotPartial(pdp4, rug = T, train = crime_train, alpha = 0.1)
```

## IML

The `iml` package implements a variety of tools for interpreting machine learning models. It requires us to first create a `Predictor()` container as a helper object. 

```{r}
X <- crime_train[which(names(crime_train) != "D_crime")]
predictor <- Predictor$new(rf, data = X, y = crime_train$D_crime, type = "prob")
```

On this basis, we can calculate feature importances based on permutations of the data. Note that more repetitions would help to stabilize results.

```{r}
imp <- FeatureImp$new(predictor, loss = "ce", n.repetitions = 2)
plot(imp)
imp$results
```

From here on we modify the `Predictor()` container and only focus on the outcome category `high` (crimes).

```{r}
predictor <- Predictor$new(rf, data = X, y = crime_train$D_crime, type = "prob",
                           class = "high")
```

The `FeatureEffect()` function allows to compute PDP, ICE and ALE plots, with ALE being the default. We create ALE plots for the same features as above with `pdp`.

```{r}
ale <- FeatureEffect$new(predictor, feature = "PctIlleg")
plot(ale)
ale$set.feature("racePctWhite")
plot(ale)
```

Global surrogate models can help to summarize rules based on a complex ML model. Here we build a small decision tree to approximate our random forest model.

```{r}
tree <- TreeSurrogate$new(predictor, maxdepth = 2)
plot(tree)
tree$r.squared
```

Finally, we can use local surrogate models such as LIME to explain predictions for individual observations. In the following we focus on the first case in our training data set.

```{r}
crime_train[1,]
predict(rf, newdata = crime_train[1,], type = "prob")
```

LIME can be used to built a simple local model for this data point. Here we restrict the number of features that we want to use in the local model to `k = 3` and print/ plot the LIME results.

```{r}
lime <- LocalModel$new(predictor, x.interest = X[1,], k = 3)
lime$results
plot(lime)
```

## References

* https://cran.r-project.org/web/packages/iml/vignettes/intro.html
