---
title: "Boosting I"
output: 
  learnr::tutorial:
    progressive: true
runtime: shiny_prerendered
tutorial:
  id: "boosting-1"
  version: 0.5
---

## Setup

```{r results='hide', message=FALSE, warning=FALSE}
library(learnr)
library(caret)
library(fastAdaboost)
library(gbm)
library(rpart)
library(pROC)
library(mlforsocialscience)
```

### Data

In this notebook we (again) use the drug consumption data. The data contains records for 1885 respondents with personality measurements (e.g. Big-5), level of education, age, gender, country of residence and ethnicity as features. In addition, information on the usage of 18 drugs is included. 

Source: https://archive.ics.uci.edu/ml/datasets/Drug+consumption+%28quantified%29

```{r}
data(drugs)
```

First, we build a dummy variable on LSD usage as our outcome of interest.

```{r}
drugs$D_LSD <- "LSD"
drugs$D_LSD[drugs$LSD == "CL0"] <- "no_LSD"
drugs$D_LSD <- as.factor(drugs$D_LSD)
drugs$D_LSD <- relevel(drugs$D_LSD, "no_LSD")
table(drugs$LSD, drugs$D_LSD)
summary(drugs$D_LSD)
```

Then we split the data into a training and a test part, using `createDataPartition` from `caret`.

```{r}
set.seed(9453)
inTrain <- createDataPartition(drugs$D_LSD, 
                               p = .8, 
                               list = FALSE, 
                               times = 1)

drugs_train <- drugs[inTrain,]
drugs_test <- drugs[-inTrain,]
```

## AdaBoost

In order to build a set of prediction models it is helpful to follow the `caret` workflow and first decide how to conduct model tuning. Here we use 5-Fold Cross-Validation, mainly to keep computation time to a minimum. `caret` offers many performance metrics, however, they are stored in different functions that need to be combined first.

```{r}
evalStats <- function(...) c(twoClassSummary(...),
                             defaultSummary(...),
                             mnLogLoss(...))
?defaultSummary()
```

Now we can specify the `trainControl` object.

```{r}
ctrl  <- trainControl(method = "cv",
                      number = 5,
                      summaryFunction = evalStats,
                      classProbs = TRUE)
```

As a first method we try out AdaBoost as implemented in the `fastAdaboost` package. Specifically, Adaboost.M1 will be used with three try-out values for the number of iterations.

```{r}
grid <- expand.grid(nIter = c(50, 100, 150),
                    method = "Adaboost.M1")
```

Now we can pass these two objects on to `train`, along with the specification of the model and the method, i.e. `adaboost`. 

```{r}
set.seed(744)
ada <- train(D_LSD ~ Age + Gender + Education + Neuroticism + Extraversion +
             Openness + Agreeableness + Conscientiousness + Impulsive + SS,
            data = drugs_train,
            method = "adaboost",
            trControl = ctrl,
            tuneGrid = grid,
            metric = "logLoss")
```

List the results of the tuning process.

```{r}
ada
```

## GBM

For Gradient Boosting as implemented by the `gbm` package, we have to take care of a number of tuning parameters. Now the `expand.grid` is helpful as it creates an object with all possible combinations of our try-out values.

```{r}
grid <- expand.grid(interaction.depth = 1:3,
                    n.trees = c(500, 750, 1000), 
                    shrinkage = c(0.05, 0.01),
                    n.minobsinnode = 10)
```

List the tuning grid...

```{r}
grid
```

...and begin the tuning process.

```{r}
set.seed(744)
gbm <- train(D_LSD ~ Age + Gender + Education + Neuroticism + Extraversion +
             Openness + Agreeableness + Conscientiousness + Impulsive + SS,
             data = drugs_train,
             method = "gbm",
             trControl = ctrl,
             tuneGrid = grid,
             metric = "logLoss",
             distribution = "bernoulli",
             verbose = FALSE)
```

Instead of just printing the results from the tuning process, we can also plot them.

```{r}
plot(gbm)
```

We can also extract single trees of the GBM ensemble.

```{r}
pretty.gbm.tree(gbm$finalModel, i.tree = 1)
pretty.gbm.tree(gbm$finalModel, i.tree = 2)
```

A quick look at feature importances.

```{r}
plot(varImp(gbm), top = 15)
```

### CART

Here we add a single tree for comparison, using CART. With `rpart2`, max tree depth is the (only) tuning parameter.

```{r}
grid <- expand.grid(maxdepth = 1:15)
```

Run the tuning process with `train()`.

```{r}
set.seed(744)
cart <- train(D_LSD ~ Age + Gender + Education + Neuroticism + Extraversion +
             Openness + Agreeableness + Conscientiousness + Impulsive + SS,
              data = drugs_train,
              method = "rpart2",
              trControl = ctrl,
              tuneGrid = grid,
              metric = "logLoss")
```

Plot and print the results.

```{r}
plot(cart)
cart
```

### Logistic regression

Finally we also add a logistic regression model. Obviously we have no tuning parameter here.

```{r}
set.seed(744)
logit <- train(D_LSD ~ Age + Gender + Education + Neuroticism + Extraversion +
             Openness + Agreeableness + Conscientiousness + Impulsive + SS,
             data = drugs_train,
             method = "glm",
             trControl = ctrl)
```

We may want to take a glimpse at the regression results.

```{r}
summary(logit)
```

## Prediction

For evaluating performance, we predict the outcome in the test set in two formats. We use `predict` for predicting class membership and probabilities based on each model, respectively.

```{r}
c_ada <- predict(ada, newdata = drugs_test)
c_gbm <- predict(gbm, newdata = drugs_test)
c_cart <- predict(cart, newdata = drugs_test)
c_logit <- predict(logit, newdata = drugs_test)

p_ada <- predict(ada, newdata = drugs_test, type = "prob")
p_gbm <- predict(gbm, newdata = drugs_test, type = "prob")
p_cart <- predict(cart, newdata = drugs_test, type = "prob")
p_logit <- predict(logit, newdata = drugs_test, type = "prob")
```

Given predicted class membership, we can use the function `postResample` in order to get a short summary of each models' performance in the test set.

```{r}
postResample(pred = c_ada, obs = drugs_test$D_LSD)
postResample(pred = c_gbm, obs = drugs_test$D_LSD)
postResample(pred = c_cart, obs = drugs_test$D_LSD)
postResample(pred = c_logit, obs = drugs_test$D_LSD)
```

Creating `ROC` objects based on predicted probabilities...

```{r}
ada_roc <- roc(drugs_test$D_LSD, p_ada$LSD)
gbm_roc <- roc(drugs_test$D_LSD, p_gbm$LSD)
cart_roc <- roc(drugs_test$D_LSD, p_cart$LSD)
logit_roc <- roc(drugs_test$D_LSD, p_logit$LSD)
```

...and plotting the ROC curves.

```{r, fig.align="center"}
ggroc(list(Adaboost = ada_roc, 
           GBM = gbm_roc, 
           CART = cart_roc, 
           Logit = logit_roc)) +
  geom_segment(aes(x = 1, xend = 0, y = 0, yend = 1), 
               color="darkgrey", linetype="dashed")
```

## References

* E. Fehrman, A. K. Muhammad, E. M. Mirkes, V. Egan and A. N. Gorban (2015). The Five Factor Model of personality and evaluation of drug consumption risk. https://arxiv.org/abs/1506.06297.
